A New Metric for the Analysis of Swarms using Potential Fields

### 2018-08-20 paper.tex
Minor revisions

\begin{abstract}
  % An important criterion in the assessment of swarm coordination algorithms is
  % the extent to which they cause unnecessary movement, i.e. movement which
  % occurs as a side effect of the algorithm rather than from the fulfilment of
  % the intended function of the swarm. 
  There are many metrics defined for the analysis of swarm coordination algorithms. These metrics are usually based upon the distances between agents, the distance between agents and a fixed point or the resultant vector that \textit{potential field} effects produce. This paper examines a distance-based metric that measures a swarms overall structure using inter-agent distances and introduces an new metric that identifies a swarm's state based upon the resultant magnitude of the vectors produced by the inter-agent interactions that create the agent distribution within a swarm's structure. The algorithms used to implement the swarming feature are based upon mathematically generated cohesion and repulsion vectors between an agent and its neighbours. In comparing and contrasting the two metrics, we find that the \textit{cohesion/repulsion} metric offers a number of advantages over the distance metric. In particular, the \textit{cohesion/repulsion} metric allows the identification of the essential characteristic of a swarm as `expanding', `stable', or `contracting'. These state cannot be identified using a distance based metric. Practical swarming applications where the new metric can be applied include area filling and reconnaissance.
\end{abstract}

\begin{keywords}
Swarming, Swarm Dynamics, Swarm Metric, Algorithms.
\end{keywords}

@@

%\section{Introduction\label{methods:SwarmStability}}
\textit{Swarming} in the animal kingdom of ants, bees, fish and birds for instance has long been studied by scientists. From these studies mathematical models and algorithms have evolved such as those produced by Reynolds~\cite{REY:87}. The models and algorithms have in turn captured the interest of computing scientists who are interested in applying them to large groups of autonomous mobile \textit{agents} (`robots'). The cooperative coordination of these agents can take many forms such as following a set path~\cite{HCS:09}, existing in a static space~\cite{EP:10, GP:02, GP:04} or foraging as a colony~\cite{HER:11, GK:07}. One of the attributes of swarms that has captured the interest of scientists is that the models and algorithms used to coordinate them are generally sets of simple rules. These simple rules cause the agents to appear to work cooperatively. These rules can also generate movement that does not contribute to the goal of the swarm, this movement is a direct result of the agents attempting to move towards an optimum position within the structure. This internal movement can be identified by analysing the changes in the inter-agent interactions. Measuring this unnecessary movement allows the effectiveness of a swarming algorithm to be evaluated. If the internal movement can be reduced then the stability of the swarm is improved and more of the swarms energy is focused on the goal. Reducing the internal movement also provides a more stable platform for the deployment of sensor arrays. Both the metrics presented in this paper allow this evaluation to be carried out when all the sensor ranges are the same however the distance metric is not able to identify the internal movement accurately when the sensor ranges of each agent vary producing irregular inter-agent distances even though the interactions at the vector level indicate that the agents are in fact stable and optimum. This paper discusses the two metrics and how they differ in their approach to identify these changes. \textit{Distance-based} metrics uses variations in the inter-agent spaces as discussed by many research teams including Navarro et al.~\cite{NIM:09}, Gazi and Passino~\cite{GP:02,GP:04}, Barnes et al.~\cite{BAF:06} and Schneider and Wildermuth~\cite{SW:03}. The new metric uses the resultant magnitudes of the agents' \textit{inter-agent vectors} that are defined by interactions of an agent with its neighbours this allows the ``degree'' of a relationship to be evaluated globally and therefore allows a swarm to be identified as expanding or contracting.

Many coordination algorithms exist to control the movement of agents within a swarm. The primary goal of these algorithms is to maintain a stable swarm structure. A stable structure is one where the agents in the swarm ``act'' as a single entity. The main focus of this paper is the use of a metric that can be applied to any coordination algorithms that use \textit{potential fields} to produce a swarming effect. To facilitate this analysis a basic swarm model is introduced and then the two metrics are applied to this model to present a comparison of the information provided by a \textit{distance-based} and the new \textit{cohesion/repulsion-based} metric. The analysis includes the identification of a swarm's ``state''. The metric is then applied to a flood filling scenario to highlight a possible use of the new metric.

The remainder of this paper is structured as follows Section~\ref{Section:Related} discusses different metrics that are currently used to analyse swarms, Section~\ref{Section:SwarmModelling} introduces a basic swarm model to produce an environment that can be analysed and evaluated using the two metrics. 
Section~\ref{Section:StabilityMagnitude} discusses the mathematical model that produces the swarming effect and how the potential fields are applied. 
Section~\ref{Section::Movement1} discusses how the model is implemented over time to produce movement. Section~\ref{Section:StabilityNullVector} discusses the effect of the model on swarm stability and the effects models have on swarm movements. 
% Section~\ref{Section:Jitter} discusses the effect the mathematical model can have upon the structure of a swarm in terms of potentially wasted energy. 
Section~\ref{Section:DistanceDynamics} discusses a distance based metric approach to swarm analysis. Section~\ref{Section:MagnitudeDynamics} introduces the new \textit{cohesion/repulsion-based} metric and how it is calculated and applied. Section~\ref{Section:ApplicationFloodFilling} compares the two metrics using a scenario of a swarm being used to fill an area. Section~\ref{Section:MagnitudeDistanceComparison} discusses the results of the experiments and Section~\ref{Section:FutureWork} discusses additional work that needs to be carried out to exploit the additional information that the new metric provides.

@@

\textit{Centroidal} analysis involves identifying the centroid of a swarm, also referred to as the \textit{centre of mass}, as shown in Figure~\ref{com} as a red cross and identified as shown in Equation~\ref{eq:Centroid1}, and calculating the distances from that position to each agent.

The \textit{centroid} of the swarm ($C_s$) is determined by taking the coordinate position of each agent and calculating the mean positions as shown in Equation~\ref{eq:Centroid1} where $b_{coord}$ is each agents coordinates and $S$ is the set of agents that make up the swarm.

\begin{equation}\label{eq:Centroid1}
C_s = \frac{1}{\card{S}}\mathlarger{\sum_{b \in S}} b_{coord}
\end{equation}

 Navarro et al. use the centroid to produce a \textit{Mean Distance from the Boundary to the Center of the Group} metric~\cite{NIM:09}. This metric takes the distance of each agent from the centroid and calculates the mean and deviation to produce their measure. Gazi and Passino use this technique when analysing the stability of a swarm based on agent distribution~\cite{GP:11}. The \textit{centroid} is also used as a reference point to determine the path of a swarm. Gazi and Passino use the centroid when analysing the stability of a swarm which is designed to maintain its position by analysing the changes in the centroids position over time~\cite{GP:04}. Gazi and Passino also use a centroid based analysis in~\cite{GP:02}~to identify directional changes is agent movements by creating a ``map'' of the potential fields of a swarm in its entirety and overlaying that to the ``centre'' of the swarm.  Barnes et. al. use the centroid to analyse the effects of using bivariate normal functions in their work on using potential fields~\cite{BAF:06, BAFVM:06}.

@@

\textit{Inter-Agent Distance} analysis is a basic mechanism for analysing a swarm it to analyse the separation of all of the agents with respect to each agents visible neighbours. This is different to using the \textit{centroidal} approach as the measure is based on agent interactions rather than a resultant structure which effects the centroid. This technique is the most popular approach to analysing a swarms structure. Gazi and Passino use a distance based analysis in~\cite{GP:04, GP:11} to identify optimum distribution of agents based upon different cohesion and repulsion applications. Wu et. al. discuss using distance-based metrics to evaluate emergent behaviours in swarm based on a Viscsek model. These include calculating the average number of neighbours for each agent, grouping agents based on numbers of neighbours (to identify clusters) and the average distance between agents~\cite{WCW:11}. Schneider and Wildermuth~\cite{SW:03} used a slightly different approach to measuring performance. They identify the difference between a simulated multi-robot environment and a physical implementation and identifying the difference in the paths of the robots based on distance variance.

\textit{Magnitude} based analysis uses the \textit{potential fields} to calculate the ``degree'' of interaction between agents. Bennet and McInnes use the resultant relationship between agents to identify equilibrium states when creating structures such as rings and controlling the direction of agents to created specific transformations~\cite{BM:09}. Wu et al. introduce the concept of a \textit{rejection potential} in an attempt to generate a kinetic energy score for a swarm~\cite{WCW:11}.

This paper builds upon the work relating to the \textit{Magnitude-based} metric analysis as discussed by Bennet and McInnes and Wu et al. by expanding the analysis to the entire swarm and converting the magnitudes to a scalar value so that the metric produces an arbitrary value that can be used to compare swarm algorithms.

@@

The concept behind the area filling is to increase both the repulsion and cohesion fields over a period of time. This increase in the field effects makes the agents increase the distance between each other, expanding the swarm as a whole. The expansion increases until, due to boundary compression, the swarm is unable to expand further. In an extreme case the expansion will result in a set of field effects that create a mesh based swarm structure rather than the desired hexagonal structure that the field effects should create. Identifying the \textit{cohesion/repulsion metric} falling below zero indicates area saturation has occurred.

>>>>

\begin{abstract}
  % An important criterion in the assessment of swarm coordination algorithms is
  % the extent to which they cause unnecessary movement, i.e. movement which
  % occurs as a side effect of the algorithm rather than from the fulfilment of
  % the intended function of the swarm. 
  There are many metrics defined for the analysis of swarm coordination algorithms. These metrics are usually based upon the distances between agents, the distance between agents and a fixed point or the resultant vector that \textit{potential field} effects produce. This paper examines a distance-based metric that measures a swarm's overall structure using inter-agent distances and introduces a new metric that identifies a swarm's state based upon the resultant magnitude of the vectors produced by the inter-agent interactions that create the agent distribution within a swarm's structure. The algorithms used to implement the swarming feature are based upon cohesion and repulsion vectors between an agent and its neighbours. In comparing and contrasting the two metrics, we find that the \textit{cohesion/repulsion} metric offers a number of advantages over the distance metric. In particular, the \textit{cohesion/repulsion} metric allows the identification of the essential characteristic of a swarm as `expanding', `stable', or `contracting'. These state cannot be identified using a distance-based metric. Practical swarming applications where the new metric can be applied advantageously include area filling and reconnaissance.
\end{abstract}

\begin{keywords}
Swarming, Swarm Dynamics, Swarm Metric.
\end{keywords}

@@

%\section{Introduction\label{methods:SwarmStability}}
\textit{Swarming} in the animal kingdom of ants, bees, fish and birds for instance has long been studied by scientists. From these studies mathematical models and algorithms have evolved such as those produced by Reynolds~\cite{REY:87}. The models and algorithms have in turn captured the interest of computing scientists who are interested in applying them to large groups of autonomous mobile \textit{agents} (`robots'). The cooperative coordination of these agents can take many forms such as following a set path~\cite{HCS:09}, existing in a static space~\cite{EP:10, GP:02, GP:04} or foraging as a colony~\cite{HER:11, GK:07}. One of the attributes of swarms that has captured the interest of scientists is that the models and algorithms used to coordinate them are generally sets of simple rules. These simple rules cause the agents to appear to work cooperatively. These rules can also generate movement that does not contribute to the goal of the swarm, this movement is a direct result of the agents attempting to move towards an optimum position within the structure. This internal movement can be identified by analysing the changes in the inter-agent interactions. Measuring this unnecessary movement allows the effectiveness of a swarming algorithm to be evaluated. If the internal movement can be reduced then the stability of the swarm is improved and more of the swarms energy is focused on the goal. Reducing the internal movement also provides a more stable platform for the deployment of sensor arrays. Both the metrics presented in this paper allow this evaluation to be carried out when all the sensor ranges are the same however the distance metric is not able to identify the internal movement accurately when the sensor ranges of each agent vary producing irregular inter-agent distances even though the interactions at the vector level indicate that the agents are in fact stable and optimum. This paper discusses the two metrics and how they differ in their approach to identify these changes. \textit{Distance-based} metrics uses variations in the inter-agent spaces as discussed by many research teams including Navarro et al.~\cite{NIM:09}, Gazi and Passino~\cite{GP:02,GP:04}, Barnes et al.~\cite{BAF:06} and Schneider and Wildermuth~\cite{SW:03}. The new metric uses the resultant magnitudes of the agents' \textit{inter-agent vectors} that are defined by interactions of an agent with its neighbours this allows the ``degree'' of a relationship to be evaluated globally and therefore allows a swarm to be identified as ``expanding'', ``stable'' or ``contracting''.

Many coordination algorithms exist to control the movement of agents within a swarm. The primary goal of these algorithms are to maintain a stable swarm structure. A stable structure is one where the agents in the swarm ``act'' as a single entity. The main focus of this paper is the use of a metric that can be applied to any coordination algorithms that use \textit{potential fields} to produce a swarming effect. Most swarming algorithms define a set of equations that calculate the repulsion and cohesion effects. These two equations are the basis of the new metric.

To facilitate the analysis of the new metric a basic swarm model is introduced and then two metrics are applied to this model to present a comparison of the information provided by a \textit{distance-based} and the new \textit{cohesion/repulsion-based} metric. The analysis includes the identification of a swarm's ``state''. The metric is then applied to a flood filling scenario to highlight a possible use of the new metric.

The remainder of this paper is structured as follows Section~\ref{Section:Related} discusses different metrics that are currently used to analyse swarms, Section~\ref{Section:SwarmModelling} introduces a basic swarm model to produce an environment that can be analysed and evaluated using the two metrics. 
Section~\ref{Section:StabilityMagnitude} discusses the mathematical model that produces the swarming effect and how the potential fields are applied. 
Section~\ref{Section::Movement1} discusses how the model is implemented over time to produce movement. Section~\ref{Section:StabilityNullVector} discusses the effect of the model on swarm stability and the effects models have on swarm movements. 
% Section~\ref{Section:Jitter} discusses the effect the mathematical model can have upon the structure of a swarm in terms of potentially wasted energy. 
Section~\ref{Section:DistanceDynamics} discusses a distance-based metric approach to swarm analysis. Section~\ref{Section:MagnitudeDynamics} introduces the new \textit{cohesion/repulsion-based} metric and how it is calculated and applied. Section~\ref{Section:ApplicationFloodFilling} compares the two metrics using a scenario of a swarm being used to fill an area. Section~\ref{Section:MagnitudeDistanceComparison} discusses the results of the experiments and Section~\ref{Section:FutureWork} discusses additional work that needs to be carried out to exploit the additional information that the new metric provides.

@@

\textit{Centroidal} analysis involves identifying the centroid of a swarm, also referred to as the \textit{centre of mass}, as shown in Figure~\ref{com} as a red cross and defined in Equation~\ref{eq:Centroid1}.

\begin{equation}\label{eq:Centroid1}
  \bar{r} = \frac{1}{\card{S}}\mathlarger{\sum_{b \in S}} r_b
\end{equation}

Here, ($\bar{r}$) denotes the position vector of the swarm's \textit{centroid}; $r_b$ is the position of each agent($b$) and $S$ is the set of agents that make up the swarm.

Navarro et al. use the centroid to produce a "Mean Distance from the Boundary to the Center of the Group" metric~\cite{NIM:09}. This metric takes the distance of each agent from the centroid and calculates the mean and deviation to produce their measure. Gazi and Passino use this technique when analysing the stability of a swarm based on agent distribution~\cite{GP:11}. The \textit{centroid} is also used as a reference point to determine the path of a swarm. Gazi and Passino use the centroid when analysing the stability of a swarm which is designed to maintain its position by analysing the changes in the centroid's position over time~\cite{GP:04}. Gazi and Passino also use a centroid-based analysis in~\cite{GP:02}~to identify directional changes in agent movement by creating a ``map'' of the potential fields of a swarm in its entirety and overlaying that to the ``centre'' of the swarm.  Barnes et. al. use the centroid to analyse the effects of using bivariate normal functions in their work on using potential fields~\cite{BAF:06, BAFVM:06}.

@@

\textit{Inter-Agent Distance} analysis is a basic mechanism for analysing a swarm in terms of separation of all of the agents with respect to each agent's visible neighbours. This is different to using the \textit{centroidal} approach as the measure is based on agent interactions rather than a resultant structure which effects the centroid. This technique is the most popular approach to analysing a swarms structure. Gazi and Passino use a distance-based analysis in~\cite{GP:04, GP:11} to identify optimum distribution of agents based upon different cohesion and repulsion applications. Wu et. al. discuss using distance-based metrics to evaluate emergent behaviours in swarm based on a Viscsek model~\cite{WCW:11}. These include calculating the average number of neighbours for each agent, grouping agents based on numbers of neighbours (to identify clusters) and the average distance between agents~\cite{WCW:11}. Schneider and Wildermuth~\cite{SW:03} use a slightly different approach to measuring performance. They identify the difference in the paths of robots in a simulated versus a physical implementation and calculate a distance variance between them.

\textit{Magnitude-based} analysis uses the \textit{potential fields} to calculate the ``degree'' of interaction between agents. Bennet and McInnes use the resultant relationship between agents to identify equilibrium states when creating structures such as rings and controlling the direction of agents to created specific transformations~\cite{BM:09}. Wu et al. introduce the concept of a \textit{rejection potential} in an attempt to generate a "kinetic energy" score for a swarm~\cite{WCW:11}.

This paper builds upon the work relating to the \textit{Magnitude-based} metric analysis as discussed by Bennet and McInnes and Wu et al. by expanding the analysis to the entire swarm, deriving a scalar value can be used in the analysis of swarm behaviour.

@@

The concept behind the area filling is to increase both the repulsion and cohesion fields over a period of time. This increase in the field effects makes the agents increase the distance between each other, expanding the swarm as a whole. The expansion increases until, due to boundary compression, the swarm is unable to expand further. In an extreme case the expansion will result in a set of field effects that create a mesh-based swarm structure rather than the desired hexagonal structure that the field effects should create. Identifying the \textit{cohesion/repulsion metric} falling below zero indicates area saturation has occurred.
